---
title: "DATA612 Project 2"
author: "Michael O'Donnell"
date: "June 14, 2019"
output: html_document
---

# Overview:
In the following R code, two recommender systems are implemented on MovieLense data
First, an Item-Based Collaborative Filtering recommender model
Second, a User-Based Collaborative Filtering recommender model
After both models are implemented, both models are evaluated
This code was written with obvious help from "Building a Recommender System with R" chapters 3 and 4

### import libraries
```{r warning=FALSE}
library(recommenderlab)
library(ggplot2)
set.seed(1)
```

### import the MovieLense data
```{r}
data(MovieLense)
MovieLense
```

### View the size of the MovieLense data
```{r}
object.size(MovieLense)
object.size(as(MovieLense, "matrix"))
```

### converting the matrix into vector to see values
```{r}
vector_ratings <- as.vector(MovieLense@data)
unique(vector_ratings)
table(vector_ratings)
```

### removing the null values and turning vector into factors
```{r}
vector_ratings <- vector_ratings[vector_ratings != 0]
vector_ratings <- factor(vector_ratings)

qplot(vector_ratings)
```

### calculating and visualizing which movies have been viewed
```{r}
views_per_movie <- colCounts(MovieLense)

table_views <- data.frame(
  movie = names(views_per_movie),
  views = views_per_movie
)

table_views <- table_views[order(table_views$views,
                                 decreasing = TRUE), ]

ggplot(table_views[1:6, ], aes(x=movie, y=views)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Number of Views of Top 6 Movies")
```

### visualizing the average movie score
```{r}
average_ratings <- colMeans(MovieLense)

qplot(average_ratings) +
  stat_bin(bins = 10) +
  ggtitle("Average movie rating")
```

### view the average ratings of only movies with 100 views minimum
```{r}
average_ratings_min100 <- average_ratings[views_per_movie >= 100]

qplot(average_ratings_min100) +
  stat_bin(bins = 10) +
  ggtitle("Average movie rating (minimum 100)")
```

### selecting only data with enough ratings and power users
```{r}
# greater than 100 views
# only accounting for users that have rated at least 50 movies
ratings_movies <- MovieLense[rowCounts(MovieLense) > 50,
                             colCounts(MovieLense) > 100]
ratings_movies

#average ratings per user
avg_ratings_user <- rowMeans(ratings_movies)
```

### normalize the user ratings to zero
```{r}
ratings_movies_normalize <- normalize(ratings_movies)
```

### splitting the data into training and testing sets
```{r}
which_train <- sample(x = c(TRUE, FALSE), size = nrow(ratings_movies),
                      replace = TRUE, prob = c(0.8, 0.2))

train <- ratings_movies[which_train, ]
test <- ratings_movies[!which_train, ]
```

### use k-fold to split the users into 5 groups
```{r}
which_set <- sample(x = 1:5, size=nrow(ratings_movies),
                    replace = TRUE)
for(i in 1:5) {
  which_train <- which_set == i
  train <- ratings_movies[which_train, ]
  test <- ratings_movies[!which_train, ]
}
```

### establishing the Item Based Collaborative Filtering recommender model
```{r}
model <- Recommender(data = train, method = "IBCF",
                     parameter = list(k=30))
model
```

### apply model onto the test set (IBCF model)
```{r}
# number of items to recommend
n_recommend <- 5

predicted <- predict(object = model, newdata = test, n = n_recommend)
predicted
```

### see the list of recommended movies for the first test user (IBCF model)
```{r}
test_user_one <- predicted@items[[1]]
test_movies_one <- predicted@itemLabels[test_user_one]
test_movies_one
```

### now, recommend movies for each user in the test set (IBCF model)
```{r}
recommender_matrix <- sapply(predicted@items, function(x){
  colnames(ratings_movies)[x]
})

recommender_matrix[, 2:4]
```

### Now, to view the most frequently recommended movies (IBCF model)
```{r}
items <- factor(table(recommender_matrix))
items <- sort(items, decreasing = TRUE)
top_items <- data.frame(names(items), items)
head(top_items)
```

### We've implemented a IBCF model
### Now, we will implement a User Based Collaborative Filtering model
(on the same data)
```{r}
model <- Recommender(data = train, method = "UBCF")
model
```

### To view some more details of this model
```{r}
names(getModel(model))
```

### apply model onto the test set (UBCF model)
```{r}
# number of items to recommend
n_recommend <- 5


predicted <- predict(object = model, newdata = test, n = n_recommend)
predicted
```

### see the list of recommended movies for the first test user (UBCF model)
```{r}
test_user_one <- predicted@items[[1]]
test_movies_one <- predicted@itemLabels[test_user_one]
test_movies_one
```

### now, recommend movies for each user in the test set (UBCF model)
```{r}
recommender_matrix <- sapply(predicted@items, function(x){
  colnames(ratings_movies)[x]
})

recommender_matrix[, 2:4]
```

### View the most frequently recommended movies (UBCF model)
```{r}
items <- factor(table(recommender_matrix))
items <- sort(items, decreasing = TRUE)
top_items <- data.frame(names(items), items)
head(top_items)
```

Since we have now implemented both IBCF and UBCF systems,
let's evaluate the models!

### First, evaluating the IBCF model
```{r}
folds <- 4
items_keep <- 15
rating_threshold <- 3

eval_sets <- evaluationScheme(data = ratings_movies, method = 
                                "cross-validation", k = folds,
                              given = items_keep, goodRating = rating_threshold)

eval_model <- "IBCF"
parameters <- NULL

eval_recommender <- Recommender(data = getData(eval_sets, "train"),
                                method = eval_model, parameter = parameters)

n_recommend <- 5

eval_predicted <- predict(object = eval_recommender, newdata =
                            getData(eval_sets, "known"), n=n_recommend,
                          type = "ratings")

eval_accuracy <- calcPredictionAccuracy(x = eval_predicted,
                                        data = getData(eval_sets, 
                                                       "unknown"),
                                        byUser = FALSE)
eval_accuracy
```

### Now, evaluating the UBCF model
```{r}
folds <- 4
items_keep <- 15
rating_threshold <- 3

eval_sets <- evaluationScheme(data = ratings_movies, method = 
                                "cross-validation", k = folds,
                              given = items_keep, goodRating = rating_threshold)

eval_model <- "UBCF"
parameters <- NULL

eval_recommender <- Recommender(data = getData(eval_sets, "train"),
                                method = eval_model, parameter = parameters)

n_recommend <- 5

eval_predicted <- predict(object = eval_recommender, newdata =
                            getData(eval_sets, "known"), n=n_recommend,
                          type = "ratings")

eval_accuracy <- calcPredictionAccuracy(x = eval_predicted,
                                        data = getData(eval_sets, 
                                                       "unknown"),
                                        byUser = FALSE)
eval_accuracy
```

# Analysis
The User-Based Collaborative Filtering recommender system outperformed the IBCF